Discussion Draft

### 2020-04-01 doc/Discussion.rmd
first shot at writing the spindle logic for reversed hetc

```{r reversed.HetC, echo=FALSE}
1c. sex differences in the SAC strength connected to conserved features of gametogenesis
- Sperm SAC strong (can detect 1 unstable bivalent) and oocyte SAC weak / 'leaky' achiasmate chromosomes pass through and higher rate of annueploidy
- leaky oocyte SAC linked to acentrosome spindle (mulitple aMTOCs) and cell volumne, these are features which differ in spermatocyttes (small, centrosome spindle, 2MTOCs)

1d. directional selection could be seen under strict SAC if the fitness landscape for 'chromosome structure' is altered via genetic architecture changes or GxE effects.
```

@@

The prediction from two of the modifier models, haploid selection and SACE are that males will generally evolve lower recombination rates overall compared to females. <cases with exceptions.? we don't have evidence or reason to predict these exception cases are held>
The two locus modifier model  does parameterize cases for male biased heterochiasmy, dependent on the stage of action for the driver (MI or MII) and the linkage status (see table 1 Brandvain and Coop).

@@

Under the spindle based model faster male specific evolution of the genome wide recombination rate is primarily driven by differences in the strength of the SAC.

@@

>>>>

```{r reversed.HetC, echo=FALSE}

#spindle selection modle points

@@

The prediction from two of the modifier models, haploid selection and SACE are that males will generally evolve lower recombination rates. <Male biased heterochiasmy predicted XX> The two locus modifier model  does parameterize cases for male biased heterochiasmy, dependent on the stage of action for the driver (MI or MII) and the linkage status (see table 1 Brandvain and Coop).

@@

Under the spindle based model faster male specific evolution of the genome wide recombination rate is primarily driven by differences in the strength of the spindle assemby checkpoint (SAC). (we won't go deeply into the molecular details (but the genes and pathways are will known, we refer the reader to [@laneKauppi2019]).

This is our logic for how a effective SAC will caus puriying selection on the bivalent / chromosome structure at meiotisis I division / reuctional division (MI).

The transition from metaphse to analphase is irreversable, so everything (all bivalents in the cell) has to be set up correctly before anaphase to prevent lower fitness later in the gametogenesis pathway. If unstable bivalents are detected on the MI spindle, the SAC delays the transition and buys time for the lose bivalent to correctly attach to the MT spindle. If this is delayed long enough, apoptosis will be triggered. If there are genetic variants which increase the number of unstable bivalents, they selected against in favor to genetic variants causing faster stabilization of bivalents on the MI spindle. Thus the SAC can lead to stabilizing selection on the chromosome structure.

It is known that the strength of the SAC varies between the forms of gametogenesis.
(via the rate of anueploidy or achiasmate bivalents in mature gametes). 
Sperm have lower anueploidy rates compared to oocytes (other lines of evidence) indcating that

the SAC in sperm is more strict compared to oocytes (which have been discribed as 'leaky'). In sperm single unpaired bivalents can trigger the SAC. While in females more oocytes progress through meiosis even with unpaired bivalents.

These differences in SAC strength have been connected to conserved features of gametogenesis; centrosome spindle and cell volumne (cite) -- which result in differing morphology of the metaphase MT spindle.

<leaky oocyte SAC linked to acentrosome spindle (mulitple aMTOCs) and cell volumne, these are features which differ in spermatocyttes (small, centrosome spindle, 2MTOCs)>

Directional selection could be the result in spermatogenesis if the fitness landscape for 'chromosome structure' is altered via genetic architecture changes or GxE effects. (in oogenesis the leaky SAC has weak selection coefficient)

@@

#blah
